Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-13 infercnv/reports/gains_and_losses.Rmd
implement infercnv_dir parameter in gains_and_losses

@@

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
infercnv_dir <- 'out/Zhang2023/by_celltype/'

# convert to sce
sce <- xfun::cache_rds({

@@

```{r plot_umaps, fig.height = 10}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
  coord_fixed()
```

@@

```{r plot_gains_and_losses_by_ct, fig.height = 20, eval = F}
# at single-cell level
unique(sce$cell_type) %>%
  purrr::walk(function(ct) {
    c('dupli', 'loss') %>%
      purrr::walk(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        p_mat <-
          p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
        p_annot <- 
          p_dat[, c('sample'), drop = F] %>%
          as.data.frame()
        p <-
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
            legend = F,
            main = paste0('celltype: ', ct, '\ndirection: ', direction))
        print(p)
      })
  })
```

@@

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_type')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction))
  })
```

#### chr7 gain

@@

##### Pseudobulk

@@

  # we assign chr7 status as >90% of genes affected by a duplication
  sce@colData$chr7_gain <-
    as.numeric(sce@colData$proportion_dupli_chr7 > 0.9)

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, sce$cell_type == 'Stromal cell'], 
      group_by = vars(sample, chr7_gain)
    )

@@

```{r test_ge}
# get differential expression
de_res <-
  glmGamPoi::test_de(
    fit, 
    contrast = 
      cond(chr7_gain = 1) -
      cond(chr7_gain = 0)
  )
```

```{r plot_de}
# minimum accepted adj_pval
min_p <- 0.05

# define up/downregulation
de_res <-
  de_res %>% 
  tibble::as_tibble() %>%
  dplyr::mutate(
    direction = dplyr::case_when(
      lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
      lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
      TRUE ~ NA
    )
  )

# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name),
    min.segment.length = 0, nudge_x = 0.5) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey')
```

>>>>

# Pre-processing

@@

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'
infercnv_dir <- params$infercnv_dir

# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

# convert obj to sce
sce <- xfun::cache_rds({

@@

```{r plot_umaps}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
  coord_fixed()
```

@@

```{r plot_gains_and_losses_by_ct}
# at single-cell level
unique(sce$cell_type) %>%
  purrr::map(function(ct) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        p_mat <-
          p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
        p_annot <- 
          p_dat[, c('sample'), drop = F] %>%
          as.data.frame()
        pheatmap::pheatmap(
          p_mat, 
          show_rownames = F, 
          annotation_row = p_annot,
          # hclust error: size cannot be NA nor exceed 65536
          cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
          legend = F,
          main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
          print()
      })
  })
```

@@

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_type')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction)) %>%
      print()
  })
```

# chr7 gain

@@

# Pseudobulking and DGEA

@@

  # we assign chr7 status as 
  # >80% of genes affected by a duplication -> chr7+
  # <20% of genes affected by a duplication -> chr7 WT
  sce@colData$chr7_gain <-
    ifelse(sce$proportion_dupli_chr7 > 0.8,
           1, ifelse(sce$proportion_dupli_chr7 < 0.2, 0, NA))

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, sce$cell_type == 'Stromal cell' & !is.na(sce$chr7_gain)], 
      group_by = vars(sample, chr7_gain)
    )

@@

```{r test_ge}
# minimum accepted adj_pval
min_p <- 0.05

# get differential expression
de_res <- xfun::cache_rds({

  # test differences
  de_res <-
    glmGamPoi::test_de(
      fit, 
      contrast = 
        cond(chr7_gain = 1) -
        cond(chr7_gain = 0)
    )

  # define significant up/downregulation and position
  de_res <-
    de_res %>% 
    tibble::as_tibble() %>%
    dplyr::mutate(
      direction = dplyr::case_when(
        lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
        lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
        TRUE ~ NA)) %>%
    dplyr::left_join(coords)

  de_res
}, file = 'de_res.rds', rerun = params$rerun)
```

```{r plot_de}
# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = log2(1.5), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```
